---
title: Exponential Distribution and Central Limit Theorem Simulation Exercise
author: "Nikhil"
date: "December 31, 2016"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    fig_width: 5
    fig_height: 3
---
```{r, include=FALSE}
echo = TRUE  # Make code always visible
options(scipen = 1)  # Turn off scientific notations for numbers
lambda = 0.2 # setting the laambda parameter as per assignment statement
noExp = 40 # number of exponentials
noSim = 1000 # 10k simulations
# Generating and plotting Exponential Distribution
dsExp = rexp (10000, lambda)

# Exponential Distribution mean calculation
vectMeans = NULL
for (i in 1:noSim) vectMeans = c (vectMeans, mean (rexp(noExp, lambda)))
```

## 1. Justification and Background
The shiny application developed explains the Effect of Lambda, No. of Observations and No. of Exponentials in Exponential Distribution and Central Limit Theorem Simulation

There are 3 slider controls to change either of the parameters. Based on the changes in slider controls, the graphs are updated automatically. Refreshing the graphs lags a bit due to the random number generation.

The reason to develop this particular app using shiny because of the following reasons

- The power of interacive charts and apps resides in the capacity to engage users to act upon controls and see immediate responses in parameters inputs and changes.

- Analytics, data science and statitical analysis benefit significantly from the new power unleashed by open source code and the internet.

```{r, cache=FALSE}
# 5.1 Generating and plotting Exponential Distribution
print (paste ("Actual and Theoretical means are", round(mean(vectMeans),2), round (1/lambda, 2)))
# 5.2 Sample Variance versus Theoretical Variance
print (paste ("Actual and Theoretical Variances are", var(vectMeans), "((1/",lambda,")^2)/",noExp,"=",(((1/lambda)^2)/noExp)))
```


## 2. Exponential Distribution - Scatter Plot and histogram
The scatter plot containing exponential distribution with certain lambda and number of observations plots the random numbers generated while performing the experiments. Similarly, the histogram "Exponential Distribution's mean distribution" graph displays the distribution of the mean of the random numbers generated.
<div class="columns-2">
```{r, echo=FALSE}
p=plot (dsExp, pch=19, cex=0.6, 
      main=paste("The exponential distribution \nwith ",
        "lambda=",lambda,"and \n10,000 observations",sep=" "),
      ylab="rexp (10000, lambda)")

q=hist (vectMeans, col="yellow", 
      main="Exponential Distribution's \nmean distribution", 
      breaks = 40, xlab="Means")
q=rug (vectMeans)
```
</div>


## 3. Normal vs. Exponential Distribution
Histogram "for "Theoretical vs. Actual mean for Exponential Distribution"" displays the difference between both the values which is very minimal confined to second digit after decimal point. Histogram "Exponential and Normal Distribution" displays the calculated distribution of means of random sampled Exponantial Distributions overlaps with the Normal Distribution, due to the Central Limit Theorem. The more samples we would get, the closer will the density distribution be to the normal distribution bell curve.
<div class="columns-2">
```{r, echo=FALSE}
r=hist(vectMeans, col="red", 
     main="Theoretical Mean vs \nActual Mean for \nExponential Distribution",
     breaks=20, xlab="Means")
r=abline(v=mean(vectMeans), lwd="4", col="green")

s=hist(vectMeans, prob=TRUE, col="lightblue", main="Exponential and Normal Distribution", breaks=40)
s=lines(density(vectMeans), lwd=3, col="red", xlab="Means")
```
</div>

##Code & other details
shinyUI(fluidPage(sidebarLayout( sidebarPanel( 
    sliderInput("lambda", "Lambda or Theorotical Mean:", min = 0.1, max = 1, value = 0.2, step = 0.1), sliderInput("noSim", "No. of Observations / Simulations:", min = 1000, max = 100000, value = 1000, step = 1000), sliderInput("noExp", "No. of Exponentials:", min = 10, max = 100, value = 40, step = 10), h5("Use the slider controls on the left pane to adjust / select:"), h5("    1. Lambda / Theorotical Mean"), h5("    2. No. of Observations / Simulations"), h5("    3. No. of Exponentials"), h5("The charts will be adjusted automatically to reflect the impact of selected inputs on the distribution")),
    mainPanel(fluidRow(column (5, plotOutput("scatterPlot")),column (5, plotOutput("histPlot1"))), fluidRow(column (5, plotOutput("histPlot2")),column (5, plotOutput("histPlot3")))))

shinyServer(function(input, output) {
  output$scatterPlot <- renderPlot({dsExp = rexp (input$noSim, input$lambda) vectMeans = NULL for (i in 1:input$noSim) vectMeans = c (vectMeans, mean (rexp(input$noExp, input$lambda))) plot (dsExp, pch=19, cex=0.6, main=paste("The exponential distribution \nwith ", "lambda=",input$lambda,"and \n",input$noSim," observations",sep=" "), ylab="rexp (noSim, lambda)")})})

Shiny application is deployed at https://nbrahman.shinyapps.io/week4_assignment/

The source code is available at 

